Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-26 bbbq_article.tex
Added feedback Geert, thanks @bogeert

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from transmembrane helices (TMHs) 
are predicted to be presented more often than expected,
based on their abundance in the proteome.
We show that the over-presentation of TMH-derived peptides is likely general, 
as it is predicted for diverse microbiota and viruses 
and for both MHC-I and MHC-II
However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show that TMHs are evolutionarily more conserved, 
as relatively less single nucleotide polymorphisms (SNPs) 
are present in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

Our immune system fights diseases and infections from pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more effective recognition of pathogens, 
are T cells which recognize peptides derived from 
exogenous antigens presented on Major Histocompatibility Complexes (MHC).

@@

Any individual's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA (Human Leukocyte Antigens) genes.
There are three genes encoding for MHC-I, which are HLA-A, HLA-B and HLA-C,
for MHC-II there are three major genes, which are HLA-DR, HLA-DQ, HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition})
[FB: is that really so?]
[RB: yes, according to the reference that is absent in the docx version to Lund et al., 2004, 'Definition of supertypes for {HLA} molecules using clustering of specificity matrices'] 
The HLA region of humans is highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA allele is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

Because of these multiple and highly polymorphic MHC genes,
the number of pathogenic peptides that can be detected is increased,
as a wide variety of MHCs (6 MHC-I and 6-8 MHC-II molecules) 
are expressed per individual
each presenting their own subset of epitopes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of very reliable prediction algorithms 
that allows for in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).

@@

Using these prediction algorithms, 
we recently showed that peptides derived 
from transmembrane helices (TMHs) 
are predicted to be more frequently presented by MHC-I 
than expected based on their abundance (\cite{bianchi2017}). 
Moreover, we showed that some well-known immunodominant peptides stem from TMHs.

This over-presentation is attributed to the fact 
that the peptide-binding groove of most MHC-I haplotypes 
is relatively hydrophobic, 
and therefore hydrophobic peptides have a higher affinity to bind
than hydrophilic ones.

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix on average 23 amino acids in length. 
TMHs are common structures in the proteins of humans and pathogens. 
Different TMH prediction tools estimate
that 15-39\% of the proteins in humans 
contain at least one TMH (\cite{ahram2006estimation}).
TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}).

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to these requirements, 
the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
[FB: Analyzing the occurrence of single nucleotide polymorphisms 
that were predicted to be inside ...][RJCB: Uh, you want to add another
paragraph here?]
Therefore, the TMHs of pathogens 
might have a lower chance to develop an escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

From an elution study, we find that epitopes derived 
from TMHs are presented on naturally occurring MHC-II,
by finding the origin of these epitopes on a reference proteome,
of which we predicted all proteins' topologies.
From in silico predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, for both MHC-I and II,
as these epitopes have a higher affinity with the presenter molecule.
From the generality of this pattern, we speculated that TMHs
are likely to be evolutionarily conserved. From an analysis using human
SNPs we found that these random point mutations are indeed less likely
to occur within TMHs. These findings encourage us to investigate the
overlooked importance of TMHs in the immune response and vaccine development.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Elution studies}\label{subsec:elution_studies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To determine if epitopes derived from TMHs are presented 
we used two elution studies, which both have sequenced
epitopes found in naturally occurring human immune cells,
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of the sequenced epitopes,
we found its possible location(s) in 
a human representative reference proteome,
with UniProt ID UP000005640\_9606.
For the epitopes that were found exactly once in the proteome,
we predicted the topology of the proteins in which the epitopes are found.
We used both TMHMM and PureseqTM to do these predictions. 
From this topology, we could determine if the epitope
overlapped with a transmembrane regions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Measuring TMH epitopes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To determine how frequently epitopes overlapping with TMHs are presented,
we applied a similar analysis strategy as described in \cite{bianchi2017}.
To investigate the generality of this, we applied this analysis
for several haplotypes of both MHC-I and MHC-II, 
as well as for a human, viral and bacterial proteome.
To summarize, for each proteome, 
all possible 9-mers (for MHC-I) or 14-mers (MHC-II) are derived. 
For each of these peptides, we determined if it overlapped with a 
TMH and if it was a binder for each haplotype.

For MHC-I, 9-mers were used, as this is the length used in our earlier
study (\cite{bianchi2017}). For MHC-II, 14-mers were used, 
as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (Mycobacterium tuberculosis (MTb), UniProt ID UP000001584) 
reference proteome. The pathogens were chosen based 
on their impact on the human 
population \richel{ref}.
We used TMHMM  \richel{ref} to predict the topology 
of the proteins within the proteome.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; \richel{ref} for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.
\richel{This is Frans' text, merge it with the current text: 
Unlike MHC-I where only one variable chain determines the landscape of the epitope binding pocket allowing for the classification of 5 HLA-A and 8 HLA-B superhaplotypes with .... variation in peptide binding repertoire, the MHC-II peptide binding groove is determined by both the alpha and beta chain. For HLA DQ both the alpha and beta chain is highly polymorphic  ..... \frans{why is HLA-DP not predicted?}
We took the alleles that represent percentage of the occurrence int he overall human population according to ... et al. 
}

This study differs in some aspects from our previous work (\cite{bianchi2017}), 
The definition of a binder differs from \cite{bianchi2017}:
in the current study a peptide is called a binder if, within a certain haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
the epitopes within a 
\emph{proteome} (see tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2}
for values), whereas the original study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
We believe our revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes

@@

To detect the evolutionary conservation of TMHs,
we collected human single nucleotide polymorphisms (SNPs)
that resulted in a single amino the substitution,
and tallied if this substitution occurred within a TMH or not.
From that tally, we determined how likely the number of SNPs
found in TMHs is by chance.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used.
Pivotal was the 'dbSNP' (\cite{sherry2001dbsnp}) database,
which contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}).
Additionally, the aptly-named 'gene' (for gene names)
and 'protein' (for proteins sequences) database were used.

@@

The first query was a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism \textbf{Homo sapiens}.
This resulted in 1077 gene IDs
% 1077 gene IDs is correct for December 2020.
% At 2021-03-01, one will get 1130 gene IDs.
% Also, one of the gene IDs that was valid back then,
% has been obsoleted.
The next query was a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database is queried for 
variations associated with the gene name.

The number of variations
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analysed, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positons of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.

Per SNP, the 'protein' NCBI database was queried for the
protein sequence.
Of each protein sequence, the protein topology was determined 
using PureseqTM.
Using the protein topology SNPs were scored to be in or outside trans membrane domains.

@@

To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
peptide elution studies were reanalyzed.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution 
studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} respectively),
for the two topology prediction tools TMHMM and 
PureseqTM (\cite{krogh2001predicting} and \cite{wang2019efficient} respectively). 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented by B cells.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MHC-I}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We recently showed that peptides derived from TMHs of human proteins 
were predicted to be over-presented on HLA-A and HLA-B. 
Here, we wondered if this finding would be general 
and if peptides stemming from predicted TMHs of microbial pathogens 
would also be over-presented on HLA-A and HLA-B.We therefor analysed all superhaplotypes of the HLA-A and HLA-B class see Figure \ref{fig:1} and found that similar as for the human proteome the percentages of MHC-I epitopes predicted to be overlapping with TMHs for our human, viral and bacterial proteome was higher as expected by TMH abundance.
See the supplementary materials (table \ref{tab:tmh_binders_mhc1}) 
The extent of the over-presentation between the various different HLA supertypes is similar for the probed proteomes, which confirms that the hydrophopbicity of the MHC binding cleft is determinant.

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{bbbq_1_smart_results/fig_f_tmh_mhc1_2_grid.png}
  \caption{
    Percentage of MHC-I epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    See table \ref{tab:tmh_binders_mhc1} for the exact TMH and epitope counts.
  }
  \label{fig:1}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MHC-II}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We next wondered if this over-representation would also generalize to MHC-II. 
Figure \ref{fig:2} shows the percentages of MHC-II epitopes overlapping 
with TMHs for our human, viral and bacterial proteome.
See the supplementary materials (table \ref{tab:tmh_binders_mhc2}) 
for the TMH and epitope counts.
We show that MHC-II has the same pattern of over-presentation
as MHC-I, yet this over-presentation is more pronounced. 
\richel{
  Showing how general our finding is (as it is explained by hydrophobicity)
  is important. See \url{https://github.com/richelbilderbeek/bbbq_article/issues/192}
}
Also for MHC-II, we observe that the extent to which TMH-derived
epitopes are presented is similar between the haplotypes, for
a human, bacterial and viral proteome.

\begin{figure}[!htbp]
  \includegraphics[height=0.8\textheight]{bbbq_1_smart_results/fig_f_tmh_mhc2_2_grid.png}
  \caption{
    Percentage of MHC-II epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    See table \ref{tab:tmh_binders_mhc2} for the exact TMH and epitope counts.
    Note that for smaller proteomes a percentage of zero is likelier.
  }
  \label{fig:2}
\end{figure}

Additionally, we investigated if MHC-I and MHC-II are equally prone
to present TMH-derived epitopes. 
\richel{Add exact method}
See \ref{fig:rel_presentation} and \ref{fig:rel_presentation_per_haplotype}.

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_conservation.png}
  \caption{
    Evolutionary conservation of SNPs in TMHs.
  }
  \label{fig:conservation}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_conservation_per_spanner.png}
  \caption{
    Evolutionary conservation of SNPs in TMHs,
    for either single- or multi-spanners
  }
  \label{fig:conservation_per_spanner}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_f_snps_found_and_expected.png}
  \caption{
    Percentage of SNPs found in TMHs.
    Each point resembles one protein, with a predicted percentage of
    TMH (x-axis) and an observed occurrence of SNPs being located
    within a TMH (y-axis).
    The dashed diagonal line shows the expected linear trend line
    if TMHs and soluble protein regions are equally conserved.
    The red line is a linear trend line that includes the membrane-associated
    proteins at the origin. 
    The blue line is a linear trend line that includes only the
    transmembrane proteins.
  }
  \label{fig:f_snps_found_and_expected}
\end{figure}

>>>>

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from human transmembrane helices (TMHs) 
are predicted to be presented more often than expected
based on their abundance in the proteome.
We show that this over-presentation of TMH-derived peptides is  general, 
as it is predicted for diverse bacteria and viruses 
and for both MHC-I and MHC-II.  However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show that TMHs are evolutionarily more conserved, 
as single nucleotide polymorphisms (SNPs) 
are present relatively less frequent in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

Our immune system fights diseases and infections from pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more effective recognition of pathogens than the innate immune response, 
are T cells which recognize peptides derived from 
exogenous antigens presented on Major Histocompatibility Complexes (MHC) class I and II.

@@

The MHC proteins are heterodimeric proteins encoded by the
HLA (Human Leukocyte Antigens) genes.
In humans, the peptide-binding groove of MHC-I is made by the alpha chain, and there are three forms of alpha chains called HLA-A, HLA-B and HLA-C. 
For MHC-II, both the alpha and the beta chains contribute to the binding groove, and there are three major forms as well, called HLA-DR, HLA-DQ and HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition})
The HLA genes of humans are highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA allele is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

While humans only express a limited number of MHC haplotypes, and therefore any individual's immune system detects only a fraction of all possible
peptide fragments, the number of pathogenic peptides that can be detected is increased at the population level, because of the multiple and highly polymorphic MHC genes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).
[GEERT: I removed the nr of MHC molecules at is is more complex, as the alpha chains coded by the maternal chromosome of HLA-DQ,DR,DP can heterodimeriza with the beta chains coded by the paternal chromosome and vice versa]

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of  prediction algorithms 
that allow for very reliable in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).[GEERT: can you put 'very reliable' in more quantitative terms (e.g. with overall more than 90\% accuracy)]

@@

Using these prediction algorithms, 
we recently predicted that peptides derived 
from transmembrane helices (TMHs) 
will be more frequently presented by MHC-I 
than expected based on their abundance (\cite{bianchi2017}). 
Moreover, we showed that some well-known immunodominant peptides stem from TMHs. This over-presentation is attributed to the fact 
that the peptide-binding groove of most MHC-I haplotypes 
is relatively hydrophobic, 
and therefore hydrophobic peptides have a higher affinity to bind
than hydrophilic ones.

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of on average 23 amino acids in length. TMHs can also be predicted from a protein sequence 
with high accuracy [GEERT: more quantitative?] by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}). 
TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate
that 15-39\% of the human proteome 
contain at least one TMH (\cite{ahram2006estimation}).

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this requirement, 
many of the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
[FB: Analyzing the occurrence of single nucleotide polymorphisms 
that were predicted to be inside ...][RJCB: Uh, you want to add another
paragraph here?][GEERT: I think it is OK like Richel wrote]
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

This study has two objectives. First, we aimed to generalize our findings by predicting the presentation of peptides from different kingdoms of life and for both MHC-I and -II. From these in silico predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, and for most haplotypes of both MHC-I and II. We confirmed this over-presentation of TMH-derived peptides by re-analyzis of a peptide elution study. Second, we tested our hypothesis that TMHs
are more evolutionarily conserved than  solvent-exposed protein regions. Our analysis of human
single nucleotide polymorphisms (SNPs) showed that random point mutations are indeed less likely
to occur within TMHs. These findings strengthen the emerging notion that TMHs are important for the T cell branch of the adaptive immune system, and hence are of  
overlooked importance in vaccine development.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Peptide elution studies}\label{subsec:elution_studies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To determine if epitopes derived from TMHs are presented 
we used two elution studies, which both have sequenced
epitopes found in naturally occurring human immune cells,
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of the sequenced epitopes,
we found its possible location(s) in 
a human representative reference proteome,
with UniProt ID UP000005640\_9606.
For the epitopes that were present only a single time in the proteome,
we predicted the topology of the proteins in which these epitopes are found.
We used both TMHMM and PureseqTM to do these predictions. 
From this topology, we could determine if the epitope
overlapped with a transmembrane region.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Predicting TMH epitopes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To predict how frequently epitopes overlapping with TMHs are presented,
we applied a similar analysis strategy as described in \cite{bianchi2017} for several haplotypes of both MHC-I and MHC-II, 
as well as for a human, viral and bacterial proteome.
To summarize, for each proteome, 
all possible 9-mers (for MHC-I) or 14-mers (MHC-II) were derived. 
For each of these peptides, we determined if it overlapped with a predicted 
TMH and if it was predicted to bind to each haplotype.

For MHC-I, 9-mers were used, as this is the length most frequently presented in MHC-I and was used in our earlier
study (\cite{bianchi2017}). For MHC-II, 14-mers were used,  as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (\emph{Mycobacterium tuberculosis} (MTb), UniProt ID UP000001584) 
reference proteome. We used TMHMM  \richel{ref} to predict the topology 
of the proteins within these proteomes.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; \richel{ref} for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The most frequent MHC-II haplotypes used  were selected, with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 MHC-II haplotypes.
\richel{This is Frans' text, merge it with the current text: 
Unlike MHC-I where only one variable chain determines the landscape of the epitope binding pocket allowing for the classification of 5 HLA-A and 8 HLA-B superhaplotypes with .... variation in peptide binding repertoire, the MHC-II peptide binding groove is determined by both the alpha and beta chain. For HLA DQ both the alpha and beta chain is highly polymorphic  ..... \frans{why is HLA-DP not predicted?}
We took the alleles that represent percentage of the occurrence int he overall human population according to ... et al. 
}

This study differs in one important aspect from our previous work (\cite{bianchi2017}), 
The definition of a binder differs from \cite{bianchi2017}:
in the current study a peptide is called a binder if, within a certain haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
the epitopes within a 
\emph{proteome} (see tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2}
for values), whereas the original study defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
We believe that our revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes

@@

To detect the evolutionary conservation of TMHs,
we collected human single nucleotide polymorphisms (SNPs)
that resulted in a single amino acid substitution,
and determined if this substitution occurred within a predicted TMH or not.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used: the 'dbSNP' (\cite{sherry2001dbsnp}) database,
which contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}), and the databases 'gene' (for gene names)
and 'protein' (for proteins sequences). [GEERT: any references for databases gene and protein?]

@@

The first query was a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism \emph{Homo sapiens}.
This resulted in 1,077 gene IDs.
% 1077 gene IDs is correct for December 2020.
% At 2021-03-01, one will get 1130 gene IDs.
% Also, one of the gene IDs that was valid back then,
% has been obsoleted.
The next query was a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database was queried for 
variations associated with the gene name.

The number of SNPs
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analysed, resulting in 38,882 SNPs.

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positons of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.[GEERT: rood practice is to present the data in a chronological fashion: thus move this to figure 1 (or put in supplement)]

Per SNP, the 'protein' NCBI database was queried for the
protein sequence.
For each protein sequence, the protein topology was determined 
using PureseqTM.
Using these predicted protein topologies, the SNPs were scored to be located within or outside trans membrane domains. [GEERT: for Table 1: Tables don't have a legend, but instead have a title and footnotes below the table.]

@@

[GEERT: I would move this subsection to below: first explain the bioinformatics predictions, and then show this as an experimental confirmation]

To obtain experimental confirmation that peptides stemming from TMHs are presented in MHC-I and MHC-II,
peptide elution studies were reanalyzed.
Table \ref{tab:elution} shows the percentages of epitopes derived 
from TMHs
found in the MHC-I (\cite{schellens2015comprehensive}) and MHC-II elution 
studies  (\cite{bergseng2015different}),
for the two topology prediction tools TMHMM (\cite{krogh2001predicting}) and PureseqTM (\cite{wang2019efficient}). 
Regardless of the prediction tool, 
at least 100 epitopes were predicted to be derived from a TMH for each condition. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented in both MHC-I and MHC-II.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Prediction of TMH-derived peptide presentation in MHC-I}
[GEERT: mooier is je subheadings samen te vatten met een conclusie. Dus: TMH-derived peptides are predicted to be over-presented in MHC-I (en dan ook voor de andere subheadings)]
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We recently predicted that peptides derived from TMHs of human proteins 
are over-presented on most haplotypes of HLA-A and HLA-B. Here, we wondered if this over-presentation is general 
and if peptides stemming from predicted TMHs of microbial pathogens 
would also be over-presented on HLA-A and HLA-B. We therefor analysed all superhaplotypes[GEERT: not explained, for instance in the Methods] of the HLA-A and HLA-B class (Figure \ref{fig:1}). Similar as for the human proteome, the percentages of MHC-I epitopes predicted to be overlapping with TMHs  was higher as expected by TMH abundance for the proteomes of SARS-CoV-2 and \emph{M. tuberculosis}.
See the supplementary materials (table \ref{tab:tmh_binders_mhc1}) [GEERT what is in supplementary table 6? The numbers?]
The extent of the over-presentation between the various different HLA supertypes is similar for the probed proteomes, which strenghtens our previous conclusion [ref] that the hydrophopbicity of the MHC-binding groove is the main factor responsible for the predicted over-presentation of TMH-derived peptides.

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{bbbq_1_smart_results/fig_f_tmh_mhc1_2_grid.png}
  \caption{
    Percentage of MHC-I epitopes predicted to be overlapping with TMHs
    for the proteomes of human, SARS-CoV-2 and \emph{M. tuberculosis}.
    The dashed horizontal lines indicate the percentage as expected by chance.
    See supplementary table \ref{tab:tmh_binders_mhc1} for the exact TMH and epitope counts.
  }
  \label{fig:1}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Prediction of TMH-derived peptide presentation in MHC-II}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We next wondered if the over-representation of TMH-derived peptided would also be present for MHC-II. 
Figure \ref{fig:2} shows the percentages of MHC-II epitopes predicted to be overlapping 
with TMHs for our human, viral and bacterial proteomes.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the TMH and epitope counts.
We predict that TMH-derived peptides are also over-presented in most MHC-II haplotypes, and this over-presentation seems even more pronounced than for MHC-I. 
\richel{
  Showing how general our finding is (as it is explained by hydrophobicity)
  is important. See \url{https://github.com/richelbilderbeek/bbbq_article/issues/192}
}
Similar to our previous results with MHC-I, we predicted that the extent to which TMH-derived
epitopes are presented by each MHC-II haplotype is similar for the human, bacterial and viral proteome.

\begin{figure}[!htbp]
  \includegraphics[height=0.8\textheight]{bbbq_1_smart_results/fig_f_tmh_mhc2_2_grid.png}
  \caption{
    Percentage of MHC-II epitopes predicted to be overlapping with TMHs
    for a human, viral and bacterial proteome.
    The dashed horizontal lines indicate the percentage as expected by chance.
    See supplementary table \ref{tab:tmh_binders_mhc2} for the exact TMH and epitope counts.
  }
  \label{fig:2}
\end{figure}

Additionally, we investigated if MHC-I and MHC-II are equally prone
to present TMH-derived epitopes. 
\richel{Add exact method}
See \ref{fig:rel_presentation} and \ref{fig:rel_presentation_per_haplotype}. [GEERT: Wat is er in 9 en 10? Wat is de conclusie? Dit moet nog wat verder uitgewerkt worden]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evolutionary conservation}
[GEERT: zie mijn punt hierbove. Een korte conclusie als titel is veel sterker. Dus iets als: TMHs are more evolutionary conserved.]
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_conservation.png}
  \caption{
    Evolutionary conservation of SNPs in TMHs. [GEERT: legenad uitwerken. Wat is precies laten zien?]
  }
  \label{fig:conservation}
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_conservation_per_spanner.png}
  \caption{
    Evolutionary conservation of SNPs in TMHs,
    for either single- or multi-spanners/ [GEERT: ook uitwerken]
  }
  \label{fig:conservation_per_spanner}
\end{figure}
